---
title: "Comparing soil nutrients between an annual burned and patch burn grazing system"
author: "Lowie Vandeplancke"
date: "2023-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Annual burning and grazing **(ABG)** has been used for cattle production to increase plant biomass for grazing but at the cost of decreasing biodiversity. Recently, patch burning and grazing **(PBG)** has become a new management method, that helps increase heterogeneity of the landscape, helping diversity, ecosystem connectivity, and stability of ecosystem function and sevices.


```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:\\Users\\louis\\OneDrive\\Documents\\UNCG\\Ecological Research')
library(tidyverse)
theme_set(theme_light())
```
Not to be included in the knit; setting working directory, loading tidyverse, and setting a light theme for graphs.

```{r, setup, include=FALSE}
PBG_n <- read.csv('PBG_N_raw_2021.csv')
PBG_p <- read.csv('PBG_P_raw_2021.csv')
PBG_treatment <- read.csv('PBG_trts.csv')
```
Info for nitrogen and phosporous samples along with the treatments.




sample data set starting with u, and dashes for Sample
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
