---
title: "Comparing Soil Nutrients Between an Annual Burned and Patch Burn Grazing System"
author: "Stephanie Mota"
date: "2023-03-13"
output: pdf_document
---
```{r, setup, include=FALSE}
library(tidyr)
library(dplyr)
```

```{r}

PBG_N_raw <- read.csv("PBG_N_compiled_raw_2021.csv")
PBG_P_raw <- read.csv("PBG_P_compiled_raw_2021.csv") %>% 
  mutate(concentration = as.character(concentration)) %>% 
  mutate(abs = as.character(abs)) %>% 
  mutate(dilution = as.character(dilution))
PBG_trts <- read.csv("PBG_trts.csv")

knitr::opts_chunk$set(echo = TRUE)
```


## Background

Cattle production often clashes with maintaining the health and biodiversity of the ecosystem. For example, annual burning and grazing (ABG) is a common practice in mesic grasslands. This increases the production of what cattle feed on at the expense of everything that lived there.
\
A newer method called patch burn grazing (PBG) is the same thing but in smaller patches. This practice was adopted with the intention to be better for the environment. The cattle graze on only the recently burned patches and have reduced grazing at other patches alone. PBG has successfully resulted in greater biodiversity compared to ABG.
\
There are no losses in doing PBG compared to ABG. 
\
To further study the effects of PBG and ABG, several resin bags were buried in the soil across the landscape. We measured soil nitrate (NO3), ammonium (NH4), and phosphorous (P) from May-August 2021. NO3 and NH4 were extracted from the resin bags by KCl extractions and P was extracted from the resin bags using HCl extractions, following standard protocols. Resin extracts were then analyzed to determine total NO3, NH4, and P in each sample. Where concentrations were too high, samples were diluted and re-measured.

```{r}
NandP <- PBG_N_raw %>% 
  bind_rows(PBG_P_raw) %>% 
  group_by(Sample) %>% 
  filter(dilution==max(dilution)) %>% 
  
  ungroup ()
```

```{r}

```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
