---
title: "Comparing soil nutrients between an annual burned and patch burn grazing system"
author: "Will Mann"
date: "2023-03-13"
output: pdf_document
---
### Cattle Production and Management

Cattle production has an incredibly detrimental impact on the surrounding wildlife. Management of cattle often *diametrically opposes* health of an environment. It is increasingly difficult to justify environmental concerns to a cattle rancher against forage for their herds.

### Annual Burning and Grazing (ABG)

ABG is commonly used by management to stimulate growth of vegetation in grazed areas. While this technique does *increase forage production*, it comes ***at the cost of habitat quality***. To move away from this management strategy, PBG has been created.

### Patch Burn Grazing (PBG)

This new burning strategy is much more localized, while still instigating growth from vegetation. This **hybrid practice allows for a mixing of values, both of conservation and business practicality**. It has been described as a *"shifting mosaic"* of vegetation, supporting environmental heterogeneity and stability. This practice is just as effective in producing forage as ABG is, with more environmental consideration.

### Soil Nutrients and Fire
**Evidence of the pros** of PBG still needs to be provided for multiple biota, as well as effects on *soil quality*.
Resin bags were placed across the landscape to discover how *ABG* and *PBG* alternatively affect soil nutrient levels of: nitrate (NO3), ammonium (NH4) and phosphorous (P). Standard HCl and KCl extraction was used for P and NO3/NH4, respectively.

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set("D:/School Stuff/Graduate Stuff/Graduate Classes/Spring 2023/UNCG_DataWrangling" = '/tmp')

setwd("D:/School Stuff/Graduate Stuff/Graduate Classes/Spring 2023/UNCG_DataWrangling")
library(tidyverse)
N21<-read.csv("PBG_N_compiled_raw_2021.csv")
P21<-read.csv("PBG_P_compiled_raw_2021.csv")
transect21<-read.csv("PBG_trts.csv") %>%
  select(-plot)
```

```{r bind, message=FALSE, warning=FALSE}
#Bind N and P together
Soil21<-rbind(N21, P21)
Samplesoil21<- Soil21 %>%
  filter(str_detect(sample, "^U")) %>%
  group_by(sample) %>%
  summarise(max = max(concentration, na.rm=TRUE)) %>%
  ungroup()
```

```{r bind, treatment and sample, warning=FALSE}
#Bind N+P to treatment
SSS<-Soil21 %>%
  select(concentration, units, test, Sample, sample) %>%
  filter(na.rm=TRUE)%>%
  filter(str_detect(sample, "^U"))

big21<-left_join(SSS,Samplesoil21)

big21[c('watershed','transect', 'plot', 'misc')]<-str_split_fixed(big21$Sample, '-', 4)
T21<-big21%>%
  select(max, concentration, sample, units, test, watershed)


STSoil21<-full_join(transect21, T21, by=c("watershed"), na.rm)
 
```

```{r treatment plot, message=FALSE, warning=FALSE}
treat<-unique(STSoil21$test)


uniquename <- STSoil21 %>% mutate(test = ifelse(test %in% c("HCl PO4_1","KCl Ammonia 10","KCL NO3_NO2 2"), c("Phosphorous","Ammonia","Nitrate"), test))



#Boxplot of concentration by treatment, colored by treatment and data
ggplot(uniquename, aes(x = test, y= as.numeric(concentration), fill = as.factor(test))) +
  geom_boxplot() +
  facet_wrap(~treatment, scales = "free") +
  scale_fill_manual(values = c("purple","red", "yellow")) +
  xlab("Nutrient") + ylab("Concentration") +
  labs(fill = "Legend")

```


```{r cv}
cv <- function(x) 100*( sd(x, na.rm = TRUE)/mean(x, na.rm = TRUE))

cvframe <- uniquename %>%
  group_by(treatment, test) %>%
  summarize(cv = cv(as.numeric(concentration))) %>%
  ungroup()

ggplot(cvframe, aes(x=test, y= as.numeric(cv), fill = as.factor(test))) +
  geom_bar(stat = "identity") +
  facet_wrap(~treatment, scales = "free") +
  scale_fill_manual(values = c("pink","orange", "green4")) +
  xlab("Nutrient") + ylab("CV of Concentration") +
  labs(fill = "Legend")

```
### ABG vs PBG

ABG *increases forage production*, it comes ***at the cost of habitat quality***. PBG is much more localized, while still instigating growth from vegetation. This **hybrid practice allows for a mixing of values, both of conservation and business practicality**. It has been described as a *"shifting mosaic"* of vegetation, supporting environmental heterogeneity and stability.

Under these considerations, with provided data, it can be said that PBG leads to less abundantly available nutrients with a lower degree of variability. To me, this seems to be taking the landscape simultaneously through multiple succession timelines all at once. Lower nutrient availability allows for slower-growing species to compete with fast growing gasses and forbs, who will likely take over the ABG plots, due to higher nutrient availability.


