---
title: "Arthropod Hebivory in the Longleaf pine savanna"
author: "Will Mann"
date: "2023-05-01"
output: pdf_document
---
###Arthropod Herbivory under Nutrient Deposition and Drought

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set("D:/School Stuff/Graduate Stuff/Graduate Classes/Spring 2023/UNCG_DataWrangling" = '/tmp')

Herb20<-read.csv("Entered Sandhills Leaf Damage Data_2020.csv", na.strings = "")

Herb20Stretch<- Herb20 %>%
  select(-Notes)%>%
  drop_na(Plot) %>%
  group_by(Block, Plot, Treatment, Replicate, Species) %>%
  gather(HerbType, HerbPerc,6:11)%>%
  ungroup() %>%
  drop_na(HerbPerc) %>%
  select(Block, Plot, Treatment, Replicate, Species, HerbType, HerbPerc) %>%
  mutate(Year="20") %>%
  filter(Species!="Chrysopsis_mariana")%>%
  filter(Species!="Lespedeza_hirta") %>%
  filter(HerbType==c("Chew","Mine.Suck"))

Herb20Div<- Herb20 %>%
  select(-Notes)%>%
  drop_na(Plot) %>%
  group_by(Block, Plot, Treatment, Replicate, Species) %>%
  gather(HerbType, HerbPerc,6:11)%>%
  ungroup() %>%
  drop_na(HerbPerc) %>%
  select(Block, Plot, Treatment, Replicate, Species, HerbType, HerbPerc) %>%
  mutate(Year=paste("20")) %>%
  mutate(Treatment_Year=paste(Treatment, "20", sep="_")) %>%
  group_by(Treatment_Year) %>%
  mutate(Totalherb=sum(HerbPerc)) %>%
  ungroup()
 

HerbCon20<- Herb20Stretch %>%
  filter(Treatment==c("C", "NPK")) %>%
  mutate(Year="20") %>%
  mutate(Treatment_Year=paste(Treatment, Year, sep="_"))



Master22<-master22


TP1<-Master22%>%
  filter(Timepoint=="1") %>%
  mutate(Year="TP1")

TP2<-Master22 %>%
  filter(Timepoint=="2") %>%
  mutate(Year="TP2")






Herb21<-read.csv("Entered Sandhills Leaf Damage Data_2021.csv", na.strings = "")

Herb21Stretch<- Herb21 %>%
  select(-Notes)%>%
  drop_na(Plot) %>%
  group_by(Block, Plot, Treatment, Rep, Species) %>%
  gather(HerbType, HerbPerc,6:11)%>%
  ungroup() %>%
  drop_na(HerbPerc) %>%
  select(-X) %>%
  mutate(Year="21") %>%
  filter(HerbType==c("Chew","Mine.Suck"))

HerbCon21<- Herb21Stretch %>%
  filter(Treatment==c("C", "NPK")) %>%
  mutate(Year="21") %>%
  mutate(Treatment_Year=paste(Treatment, Year, sep="_"))


TP1Herb22<-read.csv("Mann_Leaf_Damage_Sandhill_Gamelands_2022.csv", na.strings = "")

TP1Herb22Stretch<- TP1Herb22 %>%
  select(-Notes, -Date, -Damage.Assesser.s., -Scribe)%>%
  drop_na(Plot) %>%
  group_by(Block, Plot, Treatment, Replicate, Species) %>%
  gather(HerbType, HerbPerc,6:11)%>%
  ungroup() %>%
  drop_na(HerbPerc) %>%
  mutate(Year="TP1") %>%
  filter(HerbType==c("Chew","Mine.Suck"))

TP1HerbCon22<- TP1Herb22Stretch %>%
  filter(Treatment==c("C", "NPK")) %>%
  mutate(Year="TP1") %>%
  mutate(Treatment_Year=paste(Treatment, Year, sep="_"))


HCon20_21<-full_join(HerbCon20, HerbCon21)
HconComp<-full_join(HCon20_21, TP1HerbCon22) %>%
  filter(HerbType==c("Chew","Mine.Suck"))

HComp2020_21<- full_join(Herb20Stretch, Herb21Stretch)

####Hcompall####

HCompall<-full_join(HComp2020_21, TP1Herb22Stretch) %>%
  drop_na(HerbPerc) %>%
  filter(HerbType==c("Chew","Mine.Suck")) %>%
  group_by(Block, Plot, Treatment, Species, HerbType, Year) %>%
  summarise(avgherb=mean(HerbPerc)) %>%
  ungroup() %>%
  filter(Species!="Chrysopsis_mariana")%>%
  filter(Species!="Lespedeza_hirta") %>%
  filter(Species!="Liatris_cokeri")%>%
  filter(Species!="Dichanthelium_commutatum") %>%
  mutate(Treatment_Year=paste(Treatment, Year, sep="_"))


HCom2<-full_join(HComp2020_21, TP1Herb22Stretch) %>%
  drop_na(HerbPerc) %>%
  group_by(Block, Plot, Treatment, Species, HerbType, Year)
  summarise(avgherb=mean(HerbPerc)) %>%
  ungroup() %>%
  filter(Species!="Chrysopsis_mariana")%>%
  filter(Species!="Lespedeza_hirta") %>%
  filter(Species!="Liatris_cokeri")%>%
  filter(Species!="Dichanthelium_commutatum") %>%
  mutate(Treatment_Year=paste(Treatment, Year, sep="_"))


HerbDam<-HCompall %>%
  summarise(Plot, Species, Treatment_Year, HerbType, avgherb, Year, Treatment) %>%
  group_by(Plot, Species, Treatment_Year) %>%
  mutate(Damage= sum(avgherb))%>%
  ungroup() %>%
  mutate(totaldam=Damage) %>%
  select(Plot, Species, Treatment_Year, avgherb, totaldam, Year, Treatment, HerbType)
```

```{r bind, message=FALSE, warning=FALSE}
#Bind N and P together
Soil21<-rbind(N21, P21)
Samplesoil21<- Soil21 %>%
  filter(str_detect(sample, "^U")) %>%
  group_by(sample) %>%
  summarise(max = max(concentration, na.rm=TRUE)) %>%
  ungroup()
```

```{r bind, treatment and sample, warning=FALSE}
#Bind N+P to treatment
SSS<-Soil21 %>%
  select(concentration, units, test, Sample, sample) %>%
  filter(na.rm=TRUE)%>%
  filter(str_detect(sample, "^U"))

big21<-left_join(SSS,Samplesoil21)

big21[c('watershed','transect', 'plot', 'misc')]<-str_split_fixed(big21$Sample, '-', 4)
T21<-big21%>%
  select(max, concentration, sample, units, test, watershed)


STSoil21<-full_join(transect21, T21, by=c("watershed"), na.rm)
 
```

```{r treatment plot, message=FALSE, warning=FALSE}
treat<-unique(STSoil21$test)


uniquename <- STSoil21 %>% mutate(test = ifelse(test %in% c("HCl PO4_1","KCl Ammonia 10","KCL NO3_NO2 2"), c("Phosphorous","Ammonia","Nitrate"), test))



#Boxplot of concentration by treatment, colored by treatment and data
ggplot(uniquename, aes(x = test, y= as.numeric(concentration), fill = as.factor(test))) +
  geom_boxplot() +
  facet_wrap(~treatment, scales = "free") +
  scale_fill_manual(values = c("purple","red", "yellow")) +
  xlab("Nutrient") + ylab("Concentration") +
  labs(fill = "Legend")

```


```{r cv}
cv <- function(x) 100*( sd(x, na.rm = TRUE)/mean(x, na.rm = TRUE))

cvframe <- uniquename %>%
  group_by(treatment, test) %>%
  summarize(cv = cv(as.numeric(concentration))) %>%
  ungroup()

ggplot(cvframe, aes(x=test, y= as.numeric(cv), fill = as.factor(test))) +
  geom_bar(stat = "identity") +
  facet_wrap(~treatment, scales = "free") +
  scale_fill_manual(values = c("pink","orange", "green4")) +
  xlab("Nutrient") + ylab("CV of Concentration") +
  labs(fill = "Legend")

```


### Conclusions

ABG *increases forage production*, it comes ***at the cost of habitat quality***. PBG is much more localized, while still instigating growth from vegetation. This **hybrid practice allows for a mixing of values, both of conservation and business practicality**. It has been described as a *"shifting mosaic"* of vegetation, supporting environmental heterogeneity and stability.

Under these considerations, with provided data, it can be said that PBG leads to less abundantly available nutrients with a lower degree of variability. To me, this seems to be taking the landscape simultaneously through multiple succession timelines all at once. Lower nutrient availability allows for slower-growing species to compete with fast growing gasses and forbs, who will likely take over the ABG plots, due to higher nutrient availability.


