---
title: "**Comparing soil nutrients between an annual burned and patch burn grazing system**"
author: "Author: Carina Bravo-Chan"
subtitle: First RMarkdown
output:
  pdf_document: default
  html_document: 
    df_print: paged
    theme: paper
  word_document: default
urlcolor: blue
knitr: rmarkdown::render
---

```{r setup,warning=FALSE,include=FALSE}
    knitr::opts_knit$set(root.dir='~/Desktop/BIO_400X_R/UNCG_DataWrangling_Old')
```


## **Background:**
Cattle production often conflicts with environmental conservation and biodiversity. Temperate grasslands are managed by annual burning and grazing **(ABG)** to boost cattle fodder yield. ABG homogenizes vegetation and eliminates dead plant material, reducing habitat quality for local flora and fauna. Patch burn grazing **(PBG)** is a new management approach that promotes conservation and sustainability while maintaining cattle weight gain. Fire and grazing create a *"mosaic"* of fodder in PBG landscapes. In an unfenced unit, fire attracts grazing animals to burned areas while reducing grazing on unburned parts. Heterogeneity, caused by disturbance in a PBG landscape, increases biodiversity, connectivity, and ecosystem function and services. Crucially, PBG maintains cow output equivalent to traditional management approaches while possibly benefiting animals and environmental resources. PBG's belowground effects on soil health, carbon sequestration, bird, small mammal, insect, microbial, and plant populations are unknown.
Resin bags were buried in the soil in this area from May to August 2021 to study the effects of ABG and PBG on soil nitrate **(NO3)**, ammonium **(NH4)**, and phosphorous **(P)**. Resin extracts measured each sample's NO3, NH4, and P levels. Overconcentrations were diluted and remeasured.

### **Data:**
```{r,include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

```{r, include=FALSE ,message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(latexpdf)
```

```{r, clean data, include=FALSE}
# Read in the CSV files
N_data <- read.csv("PBG_N_compiled_raw_2021.csv")
P_data <- read.csv("PBG_P_compiled_raw_2021.csv")

# Bind the N and P files together
NP_data <- rbind(N_data,P_data) %>%
        filter(grepl("U",sample)) %>% 
        group_by(Sample,test) %>% 
        filter(dilution==max(dilution))

#Separate column Sample into four separate columns
NP_datasplit <- separate(NP_data, Sample, into = c("watershed", "transect", "plot", "col4"), sep = "-") %>% 

#Remove the 4th column created by the split above
        select(-col4)

#Join NP_datasplit & PBG_trts
PBG_trts <- read.csv("PBG_trts.csv")

#Clean PBG_trts data, split columns
PBG_combined <- merge(NP_datasplit,PBG_trts, by = c("watershed", "transect","plot"), all.x = TRUE) %>% 
  group_by(treatment) %>% 
  mutate(test=ifelse(test=="HCl PO4_1",
    "Phosphate",ifelse(test=="KCL NO3_NO2 2", "Nitrate", "Ammonia")))
```

```{r boxplot, warning=FALSE, message=FALSE, echo=FALSE}
#Create boxplot graph by Tx 
my_pal <- c("#224b5e","#94b594")
boxplot <- (ggplot(PBG_combined,aes(x=treatment,y=as.numeric(concentration),fill=treatment))+
  geom_boxplot()+
  facet_wrap(~test,scales="free_y")+
  labs(x="Treatment",y="Concentration(mg N/L)",fill="Treatment")+
  scale_fill_manual(values=my_pal)+
  ggtitle("Nutrient Concentration by Treatment Type",subtitle=waiver()))
print(boxplot)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE}
#Coefficient of variation
PBG_coeff <- PBG_combined %>% 
  group_by(treatment,test) %>% 
  summarise(conc_mean=mean(as.numeric(concentration)), conc_sd=sd(concentration), conc_coeff=conc_mean/conc_sd) %>%
  ungroup()
#Coefficient of variation - bar graph
ggplot(PBG_coeff, aes(x = test, y = conc_coeff, fill = treatment)) +
  geom_bar(stat = "identity",position="dodge") +
  labs(x = "Nutrient", y = "Concentration Coefficient") +
  scale_fill_manual(values=c("#224b5e","#94b594"))+
  ggtitle("Concentration Coefficient by Nutrient and Treatment")
```

### **Summary:**
The difference in ABG and PBG effects on soil nutrient accumulation could be attributed to the different management approaches used in each system. *ABG* involves annual burning and grazing, which can result in a more homogenized landscape and increased nitrate and ammonia accumulation in the soil. *PBG*, on the other hand, involves patch burning and grazing, resulting in a mosaic of vegetation and disturbance patterns that leads to increased phosphate accumulation in the soil. Furthermore, the belowground effects of PBG on soil health and microbial populations may contribute to nutrient accumulation differences.

Future research could look into the long-term effects of ABG and PBG on soil health and fertility, as well as their impact on aboveground biodiversity and ecosystem services, to better understand the effects of ABG and PBG on soil nutrient accumulation. Furthermore, studies could look into the possibility of combining ABG and PBG approaches in a way that balances the benefits of both systems while minimizing their negative environmental impacts. Finally, efforts could be made to disseminate the findings of these studies to cattle producers and land managers in order to encourage the adoption of more environmentally friendly and sustainable management practices.
