---
title: "Group5_RMarkdown_Answers"
author: "Arnav Pandey, Themesha Parker, Dhanush Devanand"
date: "2026-04-20"
output: pdf_document
---

--------------------------------------------------------------------------------  

# Section 2: Formatting Text Using RMarkdown Syntax  

--------------------------------------------------------------------------------

Objective: learn how to format the appearance of your document in your chosen output format.
Author: Arnav Pandey (a_pandey2@uncg.edu)   

.........................

## 2.1. Headings
.........................

Task: Read the question below, click **Knit** to render the document and return here to answer the question.  
Question: What difference do you notice about the above line that starts with
`# Section 2: Formatting a Report Using RMarkdown Syntax` compared to the line
`Objective: learn to format how your document appears in the output of your choice`?  
Answer: The font is large and bold in line starting with `# section 2.` while the other line
seems to be normal.

\

Placing a hashtag (#) at the beginning of a line turns it into a heading. 
RMarkdown supports six levels of headings. 

\


Task: Click Knit to observe how different heading levels below appear in the rendered output.

# Line 1: This is an example to show different levels of headings in RMarkdown.
## Line 2: Line 1: This is an example to show different levels of headings in RMarkdown.
### Line 3: This is an example to show different levels of headings in RMarkdown.
#### Line 4: This is an example to show different levels of headings in RMarkdown.
##### Line 5: This is an example to show different levels of headings in RMarkdown.   
###### Line 6: This is an example to show different levels of headings in RMarkdown.


Question: What happens to the size of text when you increase the number of hashtags (#)?  
Answer: Increasing numbers of hashtags decreases the font size.

\


In the example above, the higher the number of hashtags, the lower the heading level (smaller font).  

Question: How many hashtags would you use if you wanted highest level heading (largest font)?  
Answer: 1

\


Question: In the rendered output, do Line 5 and Line 6 from above example appear in the same line?  
Answer: Yes

\


Task: Go to the metadata for this assignment and change the output from `pdf_document` to `html_document`.  
Question: Did your Line 5 and Line 6 appear in different lines?  
Answer: No

\


Sometimes the way the program renders output differs between formats.
You can leave two spaces at the end of the line for a line break.  

Task: Change the output back to `pdf_document`.

Task: **Go back** to all the above example lines "Line 1" thru "Line 6" within
single set of three backticks (```       ```) to avoid conflicts with rest of the assignment.
Go back above and do as below:

```

# Line 1: This is an example to show different levels of headings in RMarkdown.
## Line 2: Line 1: This is an example to show different levels of headings in RMarkdown.
### Line 3: This is an example to show different levels of headings in RMarkdown.
#### Line 4: This is an example to show different levels of headings in RMarkdown.
##### Line 5: This is an example to show different levels of headings in RMarkdown.   
###### Line 6: This is an example to show different levels of headings in RMarkdown.

```



.........................

## 2.2 Italics
.........................  
You may sometimes want to italicize text. RMarkdown allows you to do this easily.
To make your text italics, you can enclose it within single asterisk `*[text]*`.

*Line 1: This is an example to show how to make text italics.* 

\

Task: Rewrite Line 1 below and only italicize the word "italics." Leave two spaces at the end of sentence for a line break.  
Answer: Line 1: This is an example to show how to make text *italics*.

\

You may be wondering if there is another way to make the text italics, and in fact, there is.
You can also enclose your text within single underscore `_[text]_`.

\


Task: In the paragraph below, make "Arnav," "Biological Data Wrangling and Visualization" and "RMarkdown" italics in the following paragraph using both the methods (asterisks and underscore) mentioned above at least once. 
My name is *Arnav* Pandey. I am currently enrolled in _Biological Data Wrangling and Visualization_ (BIO 457/657).
This is a task to use *RMarkdown* to make a text output italics.  

\


.........................

## 2.3 Bold
.........................  
Sometimes you may want to bold your text to emphasize its importance. RMarkdown supports that as well.  
  
Task: Read the question below, click Knit, and return to answer the question below.
Question: Which of the following lines gives you bold text?  
Line 1. This is an example to show how to make text bold.  
**Line 2: This is an example to show how to make text bold.**  
Answer:  Line 2

\

You can use two asterisks before and after the text to make it bold (`**[text]**`).   
There is also another way to make a text bold by enclosing your text within two underscores(`__[text]__`). 

Don't believe me? Knit to see if the line below is bold.  

__Line 3: This is an example to show how to make text bold.__

\

Like you practiced with italics, you can make any word or phrase within a sentence bold. See example of it below when you hit Knit.  
Line 5: This is an example to show how to make text **bold**.  
Line 4: This is an __example__ to show how to make text bold.   

\

Task: In the paragraph below, make "Arnav," "Biological Data Wrangling and Visualization" and "RMarkdown" bold in the following paragraph using
both the methods (asterisks and underscores) mentioned above at least once.   
My name is **Arnav** Pandey. I am currently enrolled in __Biological Data Wrangling and Visualization__(BIO 457/657).
This is a task to use __RMarkdown__ to make a text output bold.  

\


.........................

## 2.4 Bold and Italics
.........................

Is there a way to make a text both bold and italics at the same time? Indeed, there is!
You can enclose your text within three asterisks (`***[text]***`). or underscores (`___[text]___`).

\

Task: In the paragraph below, make "Arnav," "Biological Data Wrangling and Visualization" and "RMarkdown" bold italics using both the methods mentioned above at least once.   
My name is ***Arnav*** Pandey. I am currently enrolled in ___Biological Data Wrangling and Visualization___ (BIO 457/657).
This is a task to use ***RMarkdown*** to make a text output bold.  


\
\

In sections 2.2, 2.3, and 2.4, you learnt how to make text italics, bold, or bold+italics by enclosing the text within one, two, and three asterisks or underscores, respecitvely. You might be tempted to start enclosing a text by underscore and end it using asterisk. For example:  

*italics_  
__bold**  
***bold italics___   

Above lines will not give your desired output, so you will have to be consistent with how you enclose your texts. Use the same symbols at both ends!


.........................

## 2.5 Bulleted list
.........................

In RMarkdown, you can list your items in the RMarkdown as an unordered (bulleted) list.

To do that, you can follow these steps.
Step 1: leave an empty line before list. 
Step 2: Use '+', '-', or '*' followed by single space before you write your list item.  
Step 3: To add another item, leave two spaces at the end of the previous item and go to a new line below and repeat Step 2.  

Task: Knit to see how the items below appear in a list in the output.  

+ Apple
+ Banana 
+ Raspberry


You can also nest subitems within an item. If you want a nested list as such, just indent the subitem that you want within an item with two spaces. Below is an example of it. 

Task: Hit Knit to look at the output of the lines below:

- Apple
- Banana
- Berries 
  - Blueberries
  - Strawberries 
  - Raspberries 
  - Blackberries
  - Cranberries
- Papaya

\

Task: Copy the above list below, add an item "Melon" and nest different types of melons as its subitems.
Hit Knit to see your output.  

- Apple
- Banana
- Berries 
  - Blueberries
  - Strawberries 
  - Raspberries 
  - Blackberries
  - Cranberries
- Papaya
- Melon
  - Watermelon
  - Cantaloupe
  - Honeydew
  - Canary Melon
\



.........................

## 2.6 Numbered List
.........................

Getting a numbered list is even easier compared to unordered or bulleted list. 
To generate a numbered list, you can follow these steps.
Step 1: Leave an empty line before list. 
Step 2: Use uppercase/lowercase English alphabets, Arabic numerals, uppercase or lowercase roman numerals followed by space before you write your list item. Be consistent how you want to list.
Step 3: To add another item, leave two spaces at the end of the previous item and go to a new line below and repeat Step 2.  

I. Apple  
II. Banana  
III. Raspberry  

\

You can also nest subitems within an item. If you want a nested list as such, just indent the subitem that you want within an item with three spaces. You can use different order format for your subitems than your primary item as well. Below is an example of it. 
**Please be careful with spaces!**


1. Apple
2. Banana 
3. Berries 
   1. Blueberries
   2. Strawberries 
   3. Raspberries 
   4. Blackberries
   5. Cranberries
4. Papaya  


Note: If you want to use Roman Numerals, you may have to leave an empty line between items on list.
-------------------------------------------------------------------------------  

# Section 3: Inserting Code Chunks and Running Basic R Code  

--------------------------------------------------------------------------------
#We can insert a code chunk by clicking on the green-button with a "c" on the inside and a "+" on the upper left. 
#Click "R". 
#We can insert functions within the chunk of code. We will see this next. 

#QUESTION: What is a code chunk? (hint look back at the slides in Canvas)
Answer: A distinct section of the document that contains executable code. The chunks are executed and their results are automatically embedded into the final report
                              OR
Isolated blocks of R code that execute during the knitting process and render directly into the final document. 
#TASK: Practice inserting a code chunk. Run a simple calculation such as "2+2" below {r}. Run this code by clicking on the green side-triangle button: 
Answer:
```{r}
2+2
```

#Next, let's assign variables. Run the following variables:
x<-10
y<-5
#To view the contents of the variable x, just type 'x' and run
x

#TASK: Write a chunk of code mulitpliying the variables x and y.
Answer:
```{r}
x*y
```

#QUESTION: What was the result? 
Answer: 50
#We will analyze the built-in 'mtcars' dataset. Run the following chunk of code.

```{r}
head(mtcars)
```
#QUESTION: What do you think the head() function does?
Answer: Displays the first few rows of a data set
#To display the summary statistics for the dataset use the summary() function. Run the following chunk of code.
```{r}
summary(mtcars)
```

#Within a new chunk of code, we will code to import the entire mtcars dataset.
#Run the following code by clicking on the green side-triangle button for "Run Current Chunk"
```{r mtcars}
data(mtcars)
```

#QUESTION: After running the previous code, view the dataframe of mtcars. How many variables are in the mtcars dataset? 
Answer: 11
#Or you can view the dataset in the console by typing "mtcars" and click 'Run Selected Lines'. 
mtcars

#To find the number of rows in the dataset, we can use the nrow() function within a new chunk of code. Run the following code.
```{r}
nrow(mtcars)
```

#TASK: Write a code to find the number of columns in the 'mtcars' dataset using the ncol() function. Remember to use a new chunk of code.
Answer:
```{r}
ncol(mtcars)
```

#QUESTION: What is the result showing in your script?
Answer:11
#The 'disp' column in the dataset stands for engine displacement. Engine displacement tells you how big the engine is in cubic inches. 
#The 'mpg' column stands for fuel efficiency in miles per gallon. 
#Lets plot a graph showing fuel efficiency over displacement for each car.
#Run the following code.
plot(mtcars$disp, mtcars$mpg)

#QUESTION: What trend is the plot showing?
Answer: As engine displacement increases, the mpg (fuel efficiency/ miles per gallon) decreases 
                                    OR
As the size of the engine increases, the mpg (fuel efficiency/ miles per gallon) decreases 
                                    OR
negative correlation 
#We can also use ggplot to run this within our chunk of code.
#This permits our plot to be displayed in the final report once "knit".   
```{r}
library(ggplot2)
ggplot(mtcars, aes(x=disp, y=mpg)) + geom_point()
```

#Let's change the labels of the x-axis to 'Engine Size (cubic inches)' 
#Let's change the y-axis to 'Miles Per Gallon'
#Let's add a title called "Fuel Efficiency vs Engine Size".
```{r}
ggplot(mtcars,aes(x=disp, y=mpg))+geom_point()+
  labs(
    x="Engine Size (cubic inches)",
    y="Miles Per Gallon",
    title="Fuel Efficiency vs Engine Size"
  )
```

#We could add a legend to show which car each point corresponds to. 
#But first, we should add a column displaying the car names using dplyr.
```{r}
library(dplyr)
mtcars<-mtcars %>% 
  mutate(car_name=rownames(mtcars))
```

#Lets'bold' the title 
#Let's add a legend to differentiate each point.
```{r}
ggplot(mtcars,aes(x=disp,y=mpg, color=factor(car_name)))+
  geom_point()+
  labs(
    x="Engine Size (cubic inches)",
    y="Miles Per Gallon",
    title="Fuel Efficiency vs Engine Size",
         color="Car Types")+
  theme(plot.title = element_text(face="bold"))
```

#Lets organize the fuel efficiency from highest to lowest. Run the following.
```{r}
mtcars %>%
  select(mpg,disp) %>% 
  arrange(desc(mpg))
```

#QUESTION: Which car has the highest fuel efficiency? 
Answer: Toyota Corolla
#QUESTION: What is the size of it's engine?
Answer: 71.1 cubic inches
#TASK: Use code to create a plot comparing disp vs hp from the 'mtcars' dataset. With the following:  
#Use the 'disp' data for the x-axis and labeled 'Engine Size (cubic inches)'
#Use the 'hp' data for the y-axis and labeled 'Horsepower (HP)'
#Add an appropriate title for the plot with a 'bold' font
#Add a legend colored by 'car_name' and labeled "Car Types" 
Answer:
```{r}
ggplot(mtcars,aes(x=disp,y=hp, color=factor(car_name)))+
  geom_point()+
  labs(
    x="Engine Size (cubic inches)",
    y="Horsepower",
    title="Horsepower vs Engine Size",
         color="Car Types")+
  theme(plot.title = element_text(face="bold"))
```

#Lets organize the horsepower from highest to lowest. Run the following.
```{r}
mtcars %>%
  select(hp,disp) %>% 
  arrange(desc(hp))
```

#QUESTION: What trend is the plot showing between engine size and horsepower? 
Answer: As engine displacement (size) increases, the horsepower increases. 
                                OR 
Positive correlation between engine size and horsepower.

#QUESTION: Which cars have the highest and lowest hp?
Answer: Maserati Bora has the highest horsepower (335 hp) and Honda Civic has the lowest horsepower (52 hp).
#Don't forget to save your changes. 

----------------------------------------
------------------------------------------------------------------------
## Section 4: Add Visualizations and Tables to Your Report

### Goal

In this section, you will create graphs and a summary table using the
mtcars data set. You will also practice describing patterns that you
observe in the output.

Before beginning, make sure you can access the mtcars dataset. Run the
line below to preview the first few rows.

head(mtcars)

Also make sure you have Tidyverse installed! Run the line of code below.

library(tidyverse)

------------------------------------------------------------------------

## Part A: Create a Scatterplot

The variables wt and mpg in the mtcars data set can help us explore the
relationship between car weight and fuel efficiency.

### Task

Create a scatter plot with wt on the x-axis and mpg on the y-axis.

**Hint:** use the plot() function

```{r scatterplot}
plot(mtcars$wt, mtcars$mpg)
```

### Task

Modify your scatterplot so that it includes: a title, an x-axis label, a
y-axis label, a point color of your choice

**Hint:** Think back to previous assignments(main=, xlab= ,ylab=, and
col=)

```{r scatterplot_labeled}
plot(mtcars$wt, mtcars$mpg,
     main = "Weight vs MPG",
     xlab = "Weight (1000 lbs)",
     ylab = "Miles per Gallon",
     col = "violet")
```

### Question

What does each point in this scatterplot represent?

**Your answer:** 
Each point represents one car in the dataset and shows that car’s weight and fuel efficiency (mpg).

### Question

What relationship do you observe between weight and mpg?

**Your answer:** 
As weight increases, mpg decreases. Heavier cars tend to have lower fuel efficiency.

### Question

Does the graph suggest a positive relationship, a negative relationship,
or no relationship? Explain your answer.

**Your answer:**
The graph shows a negative relationship because the points trend downward as weight increases, meaning mpg decreases as weight increases.

------------------------------------------------------------------------

## Part B: Create a Boxplot

The variable cyl shows the number of cylinders in each car. A boxplot
can help compare fuel efficiency across cylinder groups.

### Task 

Create a boxplot of mpg by cyl using the boxplot function

```{r boxplot1}
boxplot(mpg ~ cyl, data = mtcars)
```

### Task 

Modify your boxplot so that it includes: a title, an x-axis label, a
y-axis label, and a fill color of your choice

```{r boxplot2}
boxplot(mpg ~ cyl, data = mtcars,
        main = "MPG by Number of Cylinders",
        xlab = "Number of Cylinders",
        ylab = "Miles per Gallon",
        col = "purple")
```

### Question

What does each box in this plot represent?

**Your answer:** 
Each box represents the distribution of mpg values for a specific cylinder group.

### Question

Which cylinder group appears to have the highest fuel efficiency?

**Your answer:** 
The 4-cylinder group has the highest fuel efficiency

### Question

Which cylinder group appears to have the lowest fuel efficiency?

**Your answer:** 
The 8-cylinder group has the lowest fuel efficiency.

### Question

Based on the boxplot, how does fuel efficiency seem to change as the
number of cylinders increases?

**Your answer:** 
Fuel efficiency decreases as the number of cylinders increases. Cars with more cylinders tend to have lower mpg.

------------------------------------------------------------------------

## Part C: Create a Summary Table

You will now create a table showing the average miles per gallon for
each cylinder group.

### Task 

Use dplyr to calculate the average mpg for each value of cyl. Rename the
data avg_mpg_by_cyl

**Hint:** you need to group the data by cyl then summarize the mean of
mpg.

Use the space below to write your code:

```{r summary_table, echo=TRUE}
library(tidyverse)
avg_mpg_by_cyl <- mtcars %>% 
  group_by(cyl) %>%
  summarise(avg_mpg = mean(mpg))
```
### Task 

Display your results as a clean table.

Use the space below to write your code:

```{r formatted_table}
avg_mpg_by_cyl
```

### Question

What does this table show?

**Your answer:** 
The table shows the average miles per gallon for each cylinder group, summarizing fuel efficiency across the groups.

### Question

Which cylinder group has the highest average mpg?

**Your answer:** 
The 4-cylinder group has the highest average mpg.

### Question

Which cylinder group has the lowest average mpg?

**Your answer:** 
The 8-cylinder group has the lowest average mpg.

------------------------------------------------------------------------

## Part D: Compare the Visuals

Now that you have created both graphs and the table, compare the
information they provide.

### Question 

How does the scatterplot help you understand the relationship between
weight and fuel efficiency?

**Your answer:** 
The scatterplot shows individual data points and helps visualize the relationship between weight and mpg, showing that heavier cars have lower fuel efficiency.

### Question 

How does the boxplot help you compare different groups of cars?

**Your answer:** 
The boxplot groups cars by cylinder count and shows the spread of mpg within each group, making it easier to compare fuel efficiency across different groups.

------------------------------------------------------------------------

## Part E: Now you try

1.  Use a different variable from mtcars to make an additional scatter
    plot.
2.  Change the color of your scatterplot.

Write your code below:

```{r Now_you_try}
plot(mtcars$hp, mtcars$mpg,
     main = "Horsepower vs MPG",
     xlab = "Horsepower",
     ylab = "Miles per Gallon",
     col = "darkblue")
```

### Question

What did you change, and how did it affect the output?

**Your answer:** 
I changed the x-variable from weight to horsepower and changed the color of the scatterplot. This allowed me to explore a different relationship in the data and improved the visual appearance of the graph.

------------------------------------------------------------------------

# Section 5: Improve and Finalize Your R Markdown Report

------------------------------------------------------------------------

**Objective:** Learn how to make your report more polished, readable,
and professional. In this section, we will take everything we have
learned and improve the overall quality of our R Markdown report so it
is easier to read and more useful.

Author: Dhanush Devanand

------------------------------------------------------------------------

## Step 1: Understanding Inline R Code

So far, we have been writing code in chunks. However, sometimes we want
to include results directly inside our text. **Inline R code** allows us
to insert calculations into sentences without a separate code chunk.

### Task

Copy and paste the following lines directly into your R Markdown
document, **outside** of any code chunk, and then Knit:

    The mtcars dataset has `r nrow(mtcars)` rows and `r ncol(mtcars)` columns.
    The average miles per gallon (mpg) is `r round(mean(mtcars$mpg), 2)`.

### Check

Did the numbers appear automatically in your rendered sentence?

Answer: Yes. The sentence should read: "The mtcars dataset has 32 rows and 11 columns. The average miles per gallon (mpg) is 20.09."

### Question

What is inline R code, and how is it different from a regular code
chunk?

**Answer:** Inline R code is a small piece of R code written directly inside a line of text using backticks and the letter r. Unlike a code chunk, it does not appear as a separate block — instead, it runs silently and inserts only the result into the sentence.

### Question

Why is inline R code useful when working with a dataset that might be
updated or changed over time?

**Answer:** If the dataset changes, any values written using inline R code will automatically update when the document is knitted. This means you do not have to manually rewrite numbers in your text every time the data changes.

### Task

Now try writing your own inline R code sentence. Using the mtcars
dataset, write a sentence that includes the maximum horsepower value
using inline R code. The function `max()` will be helpful here.

**Your sentence:**
The maximum horsepower in the mtcars dataset is `r max(mtcars$hp)`.

### Why This Matters

Inline R code makes your report dynamic. If the dataset changes, your
results will automatically update without you needing to rewrite
anything. This is especially important in professional or scientific
reports where accuracy matters.

------------------------------------------------------------------------

## Step 2: Improving Readability by Hiding Code

Sometimes, showing all code can make a report look cluttered. If we only
care about the output — like a graph — we can hide the code using the
`echo=FALSE` chunk option.

### Task

The chunk below currently shows the code when knitted. Add `echo=FALSE`
inside the curly braces so that only the plot appears in the output, not
the code. Then Knit your document.

```{r echo=FALSE}
plot(mtcars$wt, mtcars$mpg,
     main = "Weight vs MPG",
     xlab = "Weight (1000 lbs)",
     ylab = "Miles Per Gallon")
```

**Hint:** Your chunk header should look like this when you are done:
` ```{r echo=FALSE} `

### Check

Do you see the plot in the output but **not** the code that created it?

Answer: Yes. Only the plot should appear — the code block is hidden.

### Question

Why might you want to hide code in a final report?

**Answer:** Hiding code makes the report cleaner and easier to read, especially for an audience that only needs to see the results and does not need to follow along with the code.

### Question

Can you think of a situation where you would want to **show** your code
in a report rather than hide it? Who might be reading that report?

**Answer:** You would want to show code when sharing your work with other programmers, instructors, or collaborators who need to review, check, or reproduce your analysis.

### Task

Now apply `echo=FALSE` to the ggplot chunk you created back in Section
3. Knit the document and confirm only the plot appears.

------------------------------------------------------------------------

## Step 3: Removing Warnings and Messages

When you load packages or run certain functions, R sometimes prints
extra warnings and messages that are not important to the reader. These
can make your report look messy. You can suppress them using the chunk
options `message=FALSE` and `warning=FALSE`.

### Task — Part A: See the messages first

Run the chunk below **without** any chunk options and Knit. Read through
the output carefully and write down any extra messages or warnings you
notice below.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
summary(mtcars)
```

**What messages or warnings did you see?**
Answer: Without the chunk options, R prints messages when loading the tidyverse package, such as which packages are being attached and any conflicts with other functions.

### Task — Part B: Suppress the messages

Now add `message=FALSE` and `warning=FALSE` to the chunk header above
and Knit again.

**Hint:** Your chunk header should look like this when you are done:
` ```{r message=FALSE, warning=FALSE} `

### Check

Did the extra messages disappear from your rendered output?

Answer: Yes. The summary table still appears but the package loading messages are gone.

### Question

What is the difference between `message=FALSE` and `warning=FALSE`?
When might you need to use both at the same time?

**Answer:** `message=FALSE` suppresses informational messages such as those printed when loading a package. `warning=FALSE` suppresses warnings, which are notices that something unexpected may have happened. You might need both when loading packages that print both types of output.

### Question

Is it always a good idea to suppress warnings? Can you think of a case
where seeing a warning might actually be important?

**Answer:** No, it is not always a good idea. Warnings can alert you to real problems such as missing values, unexpected data types, or functions that are being used incorrectly. You should understand what a warning means before choosing to hide it.

### Why This Matters

Hiding unnecessary messages keeps your report clean and professional,
especially when sharing results with someone who does not need to see
behind-the-scenes R output. However, always make sure you understand
what a warning means before choosing to hide it.

------------------------------------------------------------------------

## Step 4: Hiding Code AND Output with `include=FALSE`

In Step 2, you learned that `echo=FALSE` hides the code but still shows
the output. Sometimes you may want to run a chunk silently — hiding both
the code and the output entirely. This is done with `include=FALSE`.

This is useful for setup steps like loading libraries or defining
variables that need to run but do not need to appear in the report.

### Task

Run the chunk below as-is and Knit. Note what appears in the output.
Then change the chunk option from `echo=FALSE` to `include=FALSE` and
Knit again.

```{r include=FALSE}
library(dplyr)
my_summary <- mtcars %>%
  summarize(mean_mpg = round(mean(mpg), 2),
            max_hp = max(hp),
            min_wt = min(wt))
my_summary
```

**Hint:** Your chunk header should look like this when you are done:
` ```{r include=FALSE} `

### Check

When using `include=FALSE`, does anything from this chunk appear in the
rendered output?

Answer: No. Nothing from this chunk appears in the output, but the object my_summary is still created and available to use elsewhere in the document.

### Question

What is the difference between `echo=FALSE` and `include=FALSE`?

**Answer:** `echo=FALSE` hides the code but still displays the output such as a plot or table. `include=FALSE` hides everything — both the code and the output — while still running the chunk so that any objects it creates are available for later use.

### Question

Give an example of when you would use `include=FALSE` in a real report.

**Answer:** You would use `include=FALSE` for a setup chunk at the top of your document where you load packages and define variables. The reader does not need to see this, but the code still needs to run for the rest of the report to work.

### Task

Now use the `my_summary` object created in the chunk above inside an
inline R code sentence below. Even though the chunk is hidden, the
object is still available to use. Copy and paste the lines below
**outside** of a code chunk and Knit to confirm the values appear.

    The average miles per gallon across all cars is `r my_summary$mean_mpg` and the most powerful engine produces `r my_summary$max_hp` horsepower.

### Check

Did the inline values populate correctly even though the chunk that
created them was hidden?

Answer: Yes. The sentence should read: "The average miles per gallon across all cars is 20.09 and the most powerful engine produces 335 horsepower."

------------------------------------------------------------------------

## Step 5: Controlling Figure Size

When creating plots in R Markdown, you can control the size of your
figures directly in the chunk header using `fig.width` and `fig.height`.
This helps make your report look clean and consistent.

### Task

Run the chunk below and Knit to see the default figure size.

```{r fig.width=8, fig.height=4}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "steelblue") +
  labs(title = "Weight vs Fuel Efficiency",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon") +
  theme_bw()
```

**Hint:** Your chunk header should look like this when you are done:
` ```{r fig.width=8, fig.height=4} `

### Check

Did the shape of the plot change in your rendered output?

Answer: Yes. The plot becomes wider and shorter compared to the default size.

### Question

How did changing `fig.width` and `fig.height` affect the appearance of
your plot?

**Answer:** Increasing fig.width made the plot wider and decreasing fig.height made it shorter, which changed the proportions of the plot and how spread out the points appeared.

### Question

Why might consistent figure sizing matter in a professional or academic
report?

**Answer:** Consistent figure sizes make the report look more organized and polished. Plots that are different sizes can make a report look uneven and harder to read.

### Task

Try setting `fig.width=4` and `fig.height=6` on the same chunk. Knit
and observe how the plot changes. Describe what you notice below.

**Observation:** The plot becomes narrower and taller, which compresses the points horizontally and stretches them vertically. The overall trend is still visible but the plot looks less natural compared to a wider format.

------------------------------------------------------------------------

## Step 6: Adding a Table of Contents

A table of contents makes it easier to navigate your report, especially
as it gets longer. You can add one by modifying the YAML header at the
very top of your `.Rmd` file.

### Task

Make sure the YAML header at the very top of your document looks like
this, then Knit:

```yaml
---
title: "Group5_RMarkdown"
author: "Arnav Pandey, Themesha Parker, Michelle Bautista Canuto, Dhanush Devanand"
date: "2026-04-07"
output:
  html_document:
    toc: true
---
```

### Check

Do you now see a table of contents at the top of your rendered HTML
document?

Answer: Yes. A clickable table of contents should appear at the top of the rendered HTML document listing all section headings.

### Question

How does a table of contents improve the readability of a long report?

**Answer:** A table of contents lets readers quickly jump to any section without scrolling through the entire document. This is especially helpful in longer reports with many sections.

### Question

The table of contents is automatically generated from your headings.
What does this tell you about the importance of using proper heading
levels (`#`, `##`, `###`) throughout your document?

**Answer:** It shows that using consistent and correct heading levels is important not just for visual formatting but also for the structure of the document. If headings are inconsistent, the table of contents will not accurately reflect the organization of the report.

------------------------------------------------------------------------

## Step 7: Adding Hyperlinks

You can include clickable links in your report to point readers to
additional resources. The syntax is `[link text](URL)`.

### Task

Add the following clickable link somewhere in your document and Knit:

`[R Markdown documentation](https://rmarkdown.rstudio.com)`

### Check

Does the link appear clickable after knitting?

Answer: Yes. The text should appear as a blue underlined hyperlink in the rendered output.

### Task

Now add a second hyperlink of your own. Find a webpage related to
something covered in this assignment — for example, the mtcars dataset,
ggplot2, or R Markdown chunk options — and add it below using the same
syntax.

**Your hyperlink:**
[ggplot2 documentation](https://ggplot2.tidyverse.org)

### Question

When might adding a hyperlink be useful in a data report?

**Answer:** Hyperlinks are useful when you want to direct readers to the source of your data, documentation for a package you used, or additional reading on a topic covered in the report.

------------------------------------------------------------------------

## Step 8: Writing a Strong Conclusion

Every report should end with a short conclusion that summarizes what was
done and what was found.

### Task

Write a short paragraph (3–5 sentences) below summarizing what you did
in this assignment. Consider:

- What dataset did you use?
- What visualizations or analyses did you create?
- What improvements did you make to the report in Section 5?
- Why is R Markdown useful for communicating data?

**Your conclusion:**
In this report, we used the mtcars dataset, which contains performance and design data for `r nrow(mtcars)` cars, to explore relationships between variables such as weight, fuel efficiency, engine size, and horsepower. We created scatterplots, boxplots, and summary tables to visualize and describe patterns in the data. In Section 5, we improved the report by using inline R code, hiding code and output with chunk options, controlling figure sizes, adding a table of contents, and including hyperlinks to additional resources. These techniques made the report cleaner, more dynamic, and easier to navigate. Overall, R Markdown is a powerful tool because it allows code, output, and written explanation to be combined into one professional document.

### Task

Now rewrite at least one sentence from your conclusion so that it
includes inline R code. For example, instead of writing "the dataset has
32 rows," use `` `r nrow(mtcars)` `` to generate that number
automatically.

**Your updated sentence with inline R code:**
In this report, we used the mtcars dataset, which contains performance and design data for `r nrow(mtcars)` cars, with an average fuel efficiency of `r round(mean(mtcars$mpg), 2)` miles per gallon.

------------------------------------------------------------------------

## Step 9: Final Check and Knit

Before submitting, go through the checklist below to make sure
everything is working correctly. In your `.Rmd` file, replace each `[ ]`
with `[x]` once you have confirmed that step.

- [x] Inline R code displays correctly in rendered text
- [x] Plots appear in the output without showing code (`echo=FALSE`)
- [x] A chunk runs silently using `include=FALSE`
- [x] No unnecessary warnings or messages appear (`message=FALSE`, `warning=FALSE`)
- [x] Figure sizes have been adjusted using `fig.width` and `fig.height`
- [x] Table of contents is visible at the top of the document
- [x] At least two hyperlinks are present and clickable
- [x] Conclusion paragraph is included with at least one inline R code value

Once all boxes are checked, click **Knit** one final time to generate
your completed report.