---
title: "Group5_RMarkdown"
author: "Arnav Pandey"
date: "2026-04-07"
output: pdf_document
---

--------------------------------------------------------------------------------  

# Section 2: Formatting Text Using RMarkdown Syntax  

--------------------------------------------------------------------------------

Objective: learn to format how your document appears in the output of your choice.  
Author: Arnav Pandey (a_pandey2@uncg.edu)   

.........................

## 2.1. Headings
.........................

Task: Read the question below, hit Knit to render and come back to this file to answer the question.  
Question: What did you notice about the above line that says
`# Section 2: Formatting a Report Using RMarkdown Syntax` compared to the line
`Objective: learn to format how your document appears in the output of your choice`?  
Answer:


\

Using a hashtag (#) in front of a line in RMarkdown file makes it a heading. 
RMarkdown supports six levels of heading. 

\


Task: Just render this file by hitting Knit to observe what different levels of heading below in RMarkdown looks like. 

# Line 1: This is an example to show different levels of headings in RMarkdown.
## Line 2: Line 1: This is an example to show different levels of headings in RMarkdown.
### Line 3: This is an example to show different levels of headings in RMarkdown.
#### Line 4: This is an example to show different levels of headings in RMarkdown.
##### Line 5: This is an example to show different levels of headings in RMarkdown.   
###### Line 6: This is an example to show different levels of headings in RMarkdown.


Question: What happened to the size of text when you increase the number of hashtags (#)?  
Answer: 

\


In the above line, the higher the number of hashtag the lower the heading level.  

Question: How many hashtags would you use if you wanted highest level heading (biggest font)?  
Answer: 

\


Question: Did your Line 5 and Line 6 appear in the same line?  
Answer:

\


Task: Go to the metadata for this assignment and change the output from pdf_document to html_document.  
Question: Did your Line 5 and Line 6 appear in different lines?  
Answer:

\


Sometimes how the program renders output differs between output format.
You can leave two spaces at the end of the line for a line break.  

.........................

## 2.2 Italics
.........................  
You may sometimes want your text to be italics as well. RMarkdown allows you to format a text to make it italics too. 
To make your text italics, you can enclose it within one asterisk `*[text]*`.

*Line 1: This is an example to show how to make text italics.* 

\

Task: Rewrite above line below to only italicize the word "italics." Leave two spaces at the end of sentence for a line break.  


\

You may be wondering if there is another way to make the text italics, and in fact, there is.
You can also enclose your text within one underscore `_[text]_`.

\


Task: Make "Arnav," "Biological Data Wrangling and Visualization" and "RMarkdown" italics in the following paragraph using
both the methods mentioned above at least once. 
My name is Arnav Pandey. I am currently enrolled in Biological Data Wrangling and Visualization (BIO 457/657).
This is a task to use RMarkdown to make a text output italics.  

\


.........................

## 2.3 Bold
.........................  
Sometimes you may want to bold your text to emphasize the importance of it. You can 
do that to format your text in RMarkdown as well.  
  
Task: Read the question below, hit Knit, and come back to this file to answer the question.
Question: Which of the following lines gives you bold text?  
Line 1. This is an example to show how to make text bold.  
**Line 2: This is an example to show how to make text bold.**  
Answer:  

\

You can use two asterisks before and after the text to make it bold (`**[text]**`).   
There is also another way to make a text bold by using two underscores 
before and after the text you want to make bold (`__[text]__`). 

Don't believe me? Hit knit to see if the line below is bold.  

__Line 3: This is an example to show how to make text bold.__

\

You do not have to always make the entire sentence bold every time. You can make any text within a sentence bold.
You can see example of it below when you hit Knit.  
Line 5: This is an example to show how to make text **bold**.  
Line 4: This is an __example__ to show how to make text bold.   

\

Task: Make "Arnav," "Biological Data Wrangling and Visualization" and "RMarkdown" bold in the following paragraph using
both the methods mentioned above at least once.   
My name is Arnav Pandey. I am currently enrolled in Biological Data Wrangling and Visualization (BIO 457/657).
This is a task to use RMarkdown to make a text output bold.  

\


.........................

## 2.3 Bold and Italics
.........................

Is there a way to make a text both bold and italics at the same time? Indeed, there is!
You can enclose your text within three asterisks (`***[text]***`). or underscores (`___[text]___`).

\

Task: Make "Arnav," "Biological Data Wrangling and Visualization" and "RMarkdown" bold italics in the following paragraph using both the methods mentioned above at least once.   
My name is Arnav Pandey. I am currently enrolled in Biological Data Wrangling and Visualization (BIO 457/657).
This is a task to use RMarkdown to make a text output bold.  


\
\

In sections 2.2, 2.3, and 2.4, you learnt to make texts italics, bold, or italics by enclosing the text within one, two, and three asterisks or underscores, respecitvely. You might be tempted start enclosing a text by underscore and end it using asterisk. For example:  

*italics_  
__bold**  
***bold italics___   

Above lines will not give your desired output, so you will have to be consistent with how you enclose your texts.


.........................

## 2.4 Bulleted list
.........................

In RMarkdown, you can list your items in the RMarkdown as an unordered or bulleted list.

To do that, follow these steps.
Step 1: Start by leaving an empty line before list. 
Step 2: Use '+', '-', or '*' followed by space before you write your list item.  
Step 3: To add another item, go to a new line below and repeat Step 2.  

Task: Hit Knit to see how the items below appear in a list in the output.  

+ Apple
+ Banana 
+ Raspberry


You can also nest subitems within an item. If you want a nested list as such, just indent the subitem that you want within an item with two spaces. Below is an example of it. 

Task: Hit Knit to look at the output of the lines below:

- Apple
- Banana
- Berries 
  - Blueberries
  - Strawberries 
  - Raspberries 
  - Blackberries
  - Cranberries
- Papaya

\

Task: Copy the above list below, add an item "Melon" and nest different types of melons as its subitems.
Hit Knit to see your output.  



.........................

## 2.5 Numbered List
.........................

Getting a numbered list is even easier compared to unordered or bulleted list. 
To generate a numbered list, follow these steps.
Step 1: Start by leaving an empty line before list. 
Step 2: Use uppercase or lowercase English alphabets, Arabic numerals, uppercase or lowercase roman numerals followed by space before you write your list item. Be consistent how you want to list.
Step 3: To add another item, leave two spaces at the end of the previous item and go to a new line below and repeat Step 2.  

I. Apple  
II. Banana  
III. Raspberry  

\

You can also nest subitems within an item. If you want a nested list as such, just indent the subitem that you want within an item with three spaces. You can use different order format for your subitems than your primary item as well. Below is an example of it.  

A. Apple  
B. Banana  
C. Berries  
   1. Blueberries  
   2. Strawberries   
   3. Raspberries   
   4. Blackberries  
   5. Cranberries  
D. Papaya  


Note: If you want to use Roman Numerals, you may have to leave an empty line between items on list.
------------------------------------------------------------------------
## Section 4: Add Visualizations and Tables to Your Report

### Goal

In this section, you will create graphs and a summary table using the
mtcars data set. You will also practice describing patterns that you
observe in the output.

### Instructions

Complete each task below. For each graph or table: write the code, run
the code, answer the questions in complete sentences.

Before beginning, make sure you can access the mtcars dataset. Run the
line below to preview the first few rows.

head(mtcars)

------------------------------------------------------------------------

## Part A: Create a Scatterplot

The variables wt and mpg in the mtcars data set can help us explore the
relationship between car weight and fuel efficiency.

### Task

Create a scatter plot with wt on the x-axis and mpg on the y-axis.

**Hint:** use the plot() function

```{r scatterplot}
# Write your code below this line



```

### Task

Modify your scatterplot so that it includes: a title, an x-axis label, a
y-axis label, a point color of your choice

**Hint:** Think back to previous assignments(main=, xlab= ,ylab=, and
col=)

```{r scatterplot_labeled}
# Write your updated scatterplot code below this line



```

### Question

What does each point in this scatterplot represent?

**Your answer:** [Write your response here.]

### Question

What relationship do you observe between weight and mpg?

**Your answer:** [Write your response here.]

### Question

Does the graph suggest a positive relationship, a negative relationship,
or no relationship? Explain your answer.

**Your answer:** [Write your response here.]

------------------------------------------------------------------------

## Part B: Create a Boxplot

The variable cyl shows the number of cylinders in each car. A boxplot
can help compare fuel efficiency across cylinder groups.

### Task 

Create a boxplot of mpg by cyl using the boxplot function

```{r boxplot1}
# Write your code below this line



```

### Task 

Modify your boxplot so that it includes: a title, an x-axis label, a
y-axis label, and a fill color of your choice

```{r boxplot2}
# Write your updated boxplot code below this line



```

### Question

What does each box in this plot represent?

**Your answer:** [Write your response here.]

### Question

Which cylinder group appears to have the highest fuel efficiency?

**Your answer:** [Write your response here.]

### Question

Which cylinder group appears to have the lowest fuel efficiency?

**Your answer:** [Write your response here.]

### Question

Based on the boxplot, how does fuel efficiency seem to change as the
number of cylinders increases?

**Your answer:** [Write your response here.]

------------------------------------------------------------------------

## Part C: Create a Summary Table

You will now create a table showing the average miles per gallon for
each cylinder group.

### Task 

Use dplyr to calculate the average mpg for each value of cyl. Rename the
data avg_mpg_by_cyl

**Hint:** you need to group the data by cyl then summarize the mean of
mpg.

Use the space below to write your code:

```{r summary_table}
# Write your code below this line



```

### Task 

Display your results as a clean table.

Use the space below to write your code:

```{r formatted_table}
# Write your code below this line



```

### Question

What does this table show?

**Your answer:** [Write your response here.]

### Question

Which cylinder group has the highest average mpg?

**Your answer:** [Write your response here.]

### Question

Which cylinder group has the lowest average mpg?

**Your answer:** [Write your response here.]

------------------------------------------------------------------------

## Part D: Compare the Visuals

Now that you have created both graphs and the table, compare the
information they provide.

### Question 

How does the scatterplot help you understand the relationship between
weight and fuel efficiency?

**Your answer:** [Write your response here.]

### Question 

How does the boxplot help you compare different groups of cars?

**Your answer:** [Write your response here.]

------------------------------------------------------------------------

## Part E: Now you try

1.  Use a different variable from mtcars to make an additional scatter
    plot.
2.  Change the color of your scatterplot.

Write your code below:

```{r Now_you_try}
# Write your code below this line



```

### Question

What did you change, and how did it affect the output?

**Your answer:** [Write your response here.]

------------------------------------------------------------------------



