---
title: "Independent Project"
author: "Noah Holland"
date: "2023-05-01"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### Background
|      Mite larvae of the families *Trombiculidae* and *Leeuwenhoekiidae* are parasitic in their larval stage, and are commonly referred to as "chiggers" in this parasitic stage. They are a known vector for *Orientia sp.*, the bacterial pathogen that causes the deadly disease, Scrub Typhus (Weitzel, et al. 2022). ). As well, *Rickettsia sp.* bacteria capable of causing disease in humans were detected in chigger mites collected from wild rodents in South Korea (Choi, et al. 2007), and more recently chigger mites infected with *Rickettsia felis*, *R. conorii*, and *R. typhi* were collected from rodents in North Carolina (Ponnusamy, et al. 2022). Data has been collected from seven sites in North Carolina recording both free-living and host-attached chigger mites during the summer of 2022, which were then tested for *Rickettsia sp.* bacteria as a follow-up to the 2022 Ponnusamy, et al. study. However, only a subset of that larger data set including only the number of chiggers collected and if the were free-living or host attached for each of the seven parks. While not a large data set, it will still be useful for exploring the habitat distribution of chigger mites and attack rates in different areas of the state.

### General Workflow 
```{}

```

### Packages
Before launching into the code, it's important to note what packages are being used for this data visualization project. 

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(gifski)
library(gganimate)
library(maps)
library(usmap)
library(ggmap)
library(mapdata)
```

### Code


```{R Data}
mite <- read.csv("Reuben Garshong Chigger Collection Data.csv")
```





### Literature Cited

| Weitzel, T.; Silva-de la Fuente, M.C.; Martínez-Valdebenito, C.; Stekolnikov, A.A.; Pérez R.; Vial, C.;  
|       Abarca, K.; Acosta-Jamett, G. Novel Vector of Scrub Typhus in Sub-Antarctic Chile: Evidence from 
|       Human Exposure. Clinical Infectious Diseases 2022, 74, 1862-1865. 
|       https://doi.org/10.1093/cid/ciab748 

| Choi, Y.J.; Lee, E.M.; Park, J.M.; Lee, K.M.; Han, S.H.; Kim, J.K.; Lee, S.H.; Song, H.J.; Coi, M.S.; Kim, 
|       I.S.; Park, K.H.; Jang, W.J. Molecular Detection of Various Rickettsiae in Mites (Acari: Trombiculidae) 
|       in Southern  Jeolla Province, Korea. Microbiology and Immunology. 2007, 51, 307-312. 
|       https://doi.org/10.1111/j.1348-0421.2007.tb03912.x   

| Ponnusamy, L.; Garshong, R.; McLean, B.; Wasserberg, G.; Durdeen, L.; Crossley, D.; Apperson, C.; Roe, 
|       R.M. Rickettsia felis and Other Rickettsia Species in Chigger Mites Collected from Wild Rodents in 
|       North Carolina, USA. Microorganisms 2022, 10, 1342. https://doi.org/10.3390/microorganisms10071342