---
title: "Independent project"
author: "Michelle Leyva-Espino"
date: "2026-05-04"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Project Overview: 


The dataset and research project was to determine if dogs can understand

both humans and other dogs’ emotions. This is important for domestic dogs

because humans’ express emotions differently than canines. In this

experiment, they used facial and auditory cues and timed the amount of

times the dog stared or reacted to the stimuli. 17 socialized dogs of

different breeds were used. Male and female dogs, as well as human and

canine stimuli, were used with positive and negative stimuli and

matching or mismatching vocalizations. 


The ecological questions I aim to address are if dogs can understand human

and canine emotions. How do dogs react to facial expressions that do not

match the auditory cues. Do dogs react more strongly to male or female 

stimuli, and if so, how large is that difference. In conclusion, the goal

of this experiment is to determine if dogs can understand human and canine 

emotions by the amount of time the dog stares or reacts to the experiment. 

The longer a dog stares, the more it is confirmed that the dogs understand

the emotions being shown.


# 2. Project Workflow:


1. First I will load in the packages need to clean data and run code. 

-tidyverse

-readxl 

-ggplot 2 

-dply 

-markdown 

-knitr

2. Next, I will import the dataset using readxl and save the dataset as an object title Emotions_Dogs. I will then inspect the data. 

3. Clean and tidy the dataset by removing the following columns. 

- Trial 
- Sequence 
- Observer 
- Notes 
- Date 
- Control 
- Positive/Negative (Represent in Valence column)

I will keep the rest of the columns which will be used to create our digarams. 

4. I will check for missing values and remove any missing values. 

- I will check side_con, matching and index columns. 
- I will convert columns in factors to make the data cleaner. 

5. Create new variables 

- I will create a new column with left and right interaction time together for a cleaner data set. 
- I will also create abs_index variable to see the strength of the response. 

6. Next I will test the differences by comparing

- mean valence 
- mean index 
- mean total_ time 

7. Lastly I will create figures to answers the questions in project overview section. 

Figure 1: Human vs Canine Emotional Reaction

- I will use a boxplot using valance and spp_sti columns

Figure 2: Matching vs Mismatching Stimuli 

- I will use a boxplot using valance and matching columns 

Figure 3: Male vs Female Stimuli 

- I will use a Boxplot using valance and sex_sti in order to determine the difference in dogs reactions between females and males. 

Figure 4: Strength of Response  

- I will use a barplot using mean of abs_index and sex_sti to see the strength of the responses. 

# 3. R Script: 

 loading packages 
 
```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(rmarkdown) 
library(knitr) 
```
 
 Importing dataset 
 
```{r}
emotional_dogs <- read_excel("Data+Sheet 2.xlsx")
```
 
Inspecting dataset 

```{r}
head(emotional_dogs)
str(emotional_dogs)
summary(emotional_dogs)
```
 
 Cleaning dataseet
 
```{r}
clean_emotional_dogs <- emotional_dogs %>%
  select(
    dog,
    sex,
    spp_sti,
    sex_sti,
    matching,
    condit,
    group,
    time_le,
    time_ri,
    index,
    valence
  ) %>%
  na.omit()
```

Creating new variables

```{r}
clean_emotional_dogs <- clean_emotional_dogs %>%
  mutate(
    total_time = time_le + time_ri,

    preference = case_when(
      time_le > time_ri ~ "Left",
      time_ri > time_le ~ "Right",
      TRUE ~ "Equal"
    ),

    abs_index = abs(index)
  )
```

Figure 1: Human vs Canine Emotional Reaction

```{r}
ggplot(clean_emotional_dogs, aes(x = spp_sti, y = valence)) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.3) +
  labs(
    title = "Human vs Canine Emotional Reactions",
    x = "Stimulus Type",
    y = "Valence"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11),
    panel.grid.major = element_line(color = adjustcolor("pink", alpha.f = 0.2)),
    panel.background = element_rect(fill = "lavenderblush")
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = "(1)",
    hjust = 1.2,
    vjust = 1.5,
    size = 5
  ) 
```

**Figure 1** This figure shows the difference in human vs canine emotional reactions. The dogs reacted a little more to human stimuli but the difference is slightly noticeable. 

Figure 2: Matching vs Mismatching Stimuli

```{r}
clean_emotional_dogs <- clean_emotional_dogs %>%
  mutate(
    matching = as.factor(matching)
  )

clean_emotional_dogs <- clean_emotional_dogs %>%
  mutate(
    matching = trimws(as.character(matching)),
    matching = case_when(
      matching == "0" ~ "Mismatching",
      matching == "1" ~ "Matching",
      TRUE ~ NA_character_
    ),
    matching = factor(matching, levels = c("Mismatching", "Matching"))
  )

ggplot(clean_emotional_dogs, aes(x = matching, y = index)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.alpha = 0.3) +
  labs(
    title = "Matching vs Mismatching Stimuli",
    x = "Cue Condition",
    y = "Preference Index"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11),
    panel.grid.major = element_line(color = adjustcolor("pink", alpha.f = 0.2)),
    panel.background = element_rect(fill = "lavenderblush")
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = "(2)",
    hjust = 1.2,
    vjust = 1.5,
    size = 5
  ) 
```

***Figure 2*** This figure shows 

Figure 3: Male vs Female Stimuli 

```{r}
ggplot(clean_emotional_dogs, aes(x = sex_sti, y = valence)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  stat_summary(fun = mean, geom = "line", aes(group = 1)) +
  labs(
    title = "Emotional Response: Male vs Female Stimuli",
    x = "Stimulus Sex",
    y = "Mean Valence"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11),
    panel.grid.major = element_line(color = adjustcolor("pink", alpha.f = 0.2)),
    panel.background = element_rect(fill = "lavenderblush")
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = "(3)",
    hjust = 1.2,
    vjust = 1.5,
    size = 5
  ) 
```

Figure 4: Strength of response by Simulus Sex

```{r}
ggplot(clean_emotional_dogs, aes(x = sex_sti, y = abs_index)) +
  stat_summary(fun = mean, geom = "col", alpha = 0.7) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  labs(
    title = "Strength of Response by Stimulus Sex",
    x = "Stimulus Sex",
    y = "Mean Absolute Index"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11),
    panel.grid.major = element_line(color = adjustcolor("pink", alpha.f = 0.2)),
    panel.background = element_rect(fill = "lavenderblush")
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = "(4)",
    hjust = 1.2,
    vjust = 1.5,
    size = 5
  ) 
```



 















