---
title: "Individual Project"
author: Amaliya Brown Doyoyo
date: "2026-04-27"
output: 
  pdf_document:
    toc: true
---

\pagebreak

# Overview

West Nile Virus (WNV) is an enveloped, single-stranded, RNA virus that causes West Nile fever; an infection that can range from mild, flu-like symptoms to life-threatening central nervous system effects. Mosquitoes, particularly those of the Culex species, are the primary vectors for WNV, with birds acting as the primary reservoir. WNV is considered a tropical disease; however, rising global temperatures have coincided with its increased incidence in the northern hemisphere. Because humans are almost exclusively infected with WNV through mosquito bites, ecological factors play an important role in the epidemiology of the disease. Environmental conditions such as temperature and precipitation can determine mosquito abundance in certain regions, while human demographic factors, like population density and proximity to travel infrastructure like railways, influence the frequency of vector-human contact and the spread of the virus to new regions.  

My project will use data from the 2024 dataset “Database of present and future situation of West Nile virus in the Afro-Palaearctic Pathogeographic System,” to compare the relationships between climate variables (temperature and precipitation) and population demographics (population density, access to travel) to WNV incidence across both afrotropic (Africa) and palaearctic (Europe and Asia) regions. I hypothesize that environmental variables will have a stronger association with WNV incidence in both regions.

# Pseudocode/Workflow

**I will be using the packages tidyverse, ggplot2, sf, and rnaturalearth.**



**Part One:** *I will tidy the data (making the afro_clean and pala_clean dataframes) before working on my figures.*

1. Create a ‘kelvin_to_celsius’ function to use on the temperature values in the dataset.
2. Tidy both dataframes using a pipe with the select (to select the columns of interest: mean temp, max temp, min temp, precipitation,   population density, distance to cities, distance to railroads, distance to roads), rename (renaming columns), mutate with         ‘kelvin_to_celcius’ (on the temperature columns), mutate with as.numeric (on all non-numeric column), drop_na, and mutate(Region = “”)  (changing the values in the Region column to “Afrotropical” or “Palaearctic”) functions. <br>
  
  
**Part Two:** *My first figure will be a bar graph comparing the correlation between select ecological and population factors on WNV incidence in both AFR and PAL regions.*

1. Make a wnv_combined dataframe using the bindrows function on afro_clean and pala_clean.
2. Make a wnv_corr dataframe from the wnv_combined dataframe. Group by Region. Use the summarize function to make new correlation columns using the cor command with “complete.obs” (to delete any missing values) to find the correlation between each factor and WNV frequency. Use pivot longer on all columns (except for Region) to turn the variable columns into rows that correspond to their respective correlation in a new “Correlation” column. Use the mutate function to create a “Type” column assigning each variable to “Ecological” or “Population”. 
3. Make the bar graphs using ggplot with wnv_corr. Set the x axis as “variable”, using the reorder argument to list them in ascending order based on correlation value. Set the y axis as correlation, filling with ‘Region’ for optimal comparison. Use geom_col to make the bar graphs, and coord_flip to make them horizontal for optimal virtualization. Use facet_wrap to generate two bar graphs for both ecological and population data. Change the appearance of the figure manually with theme functions. 
  
  
**Part Three:** *Make a map of the average WNV frequency for all ecological/population variables for both AFR and PAL regions. Uses Z-score to standardise the data for clearer comparison.*

1. Make the wnv_avg_corr dataframe from the wnv_corr dataframe. Group by Region and Type, then use the summarize function to find the mean correlation for each Region and Type. Use the mutate function to create a column for Z-scores. 
2. Make a wnv_continent_corr dataframe from the wnv_avg_corr dataframe. Use the mutate function to assign AFR and PAL to their respective continents with the case_when function. Make an asia_data dataframe from the wnv_avg_corr dataframe with the filter and mutate to manually assign the PAL values to the asian continent. Update the wnv_continent_corr dataframe by binding the original dataframe with the asia_data dataframe.
3. Create a world dataframe using the ne_countries functions from the sf package. Create a map_data dataframe for the plot by using left_join to join the wnv_continent_corr data with the world data by continent, using the “many-to-many” argument since both dataframes are matching several rows of data. Use the filter function to remove any countries outside of the regions of interest. 
4. Use ggplot with map_data to create a map. Use geom_sf to fill the map with zscore data and facet_wrap to make two maps for ecological and population data. Change the appearance of the figure manually with theme functions. 
  
  
**Part Four:** *Make pie charts comparing the estimated contribution of each variable on WNV in each region and in both regions combined.*

1. Create  a pie_normalized dataframe from wnv_corr that uses the mutate function to make a correlation column including squared correlation values (to make positive); group by Region and use the mutate function to make a percentage column that shows the estimated percent contribution of each factor (contribution of single factor / sum of all factors). 
2. Create a combined_pie dataset from the pie_normalized dataset; group by variable and use the summarize function to find the mean percentage for each variable. Use the mutate function to change the “Region” values to “Both.” 
3. Create the final_pie_data dataframe using the bindrows function on pie_normalized and combined_pie to combine both datasets. 
4. Use ggplot on final_pie_data with y as estimated percentage filled with variable values for optimal identification. Use geom_col and coord_polar to make the pie charts. Use facet_wraps to make three pie charts for each region. Change the appearance of the figure manually with theme functions. 

\pagebreak

# Part One: Set up

```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE, fig.width = 12, fig.height = 7, out.width = NULL, dpi = 300}
# -----------------------------------------------------------------------
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##### Comparing the Effects of Ecological vs Population Factors on West 
#### Nile Virus Incidence in Afrotropical vs Palaearctic Regions 
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# -----------------------------------------------------------------------

  # I. Preparing Workspace
  library(tidyverse)
  library(ggplot2)
  library(sf)
  library(rnaturalearth)
  
  # II. Downloading Data
  afrotropical <- read.csv("Afrotropical_region.csv")
  palaearctic <- read.csv("Palaearctic_region.csv")
  
# ---------------------------
# PART ONE: Tidying the Data
# ---------------------------

  # 1.0: kelvin_to_celsius converts Kelvins * 10 to Celsius.
  
  kelvin_to_celsius <- function(temp_k) 
  {temp_c <- (temp_k/10) - 273.15
  return(temp_c)
  }
  
  # 1.1: Making tidy afrotropical and palaearctic dataframes.

  # Tidy afrotropical dataframe
  afro_clean <- afrotropical %>% 
    select(BIO1, BIO5_MEAN, BIO6_MEAN, BIO12, DENS_POB, 
           DISTCENPOB, DISRT_RAIL, DIST_ROAD, FWNVetio, Region) %>% 
    rename(Temp_Mean = BIO1, Temp_Max = BIO5_MEAN, 
           Temp_Min = BIO6_MEAN,
           Annual_Precipitation = BIO12, Population_Density = DENS_POB, 
           Distance_Cities = DISTCENPOB,
           Distance_Railroals = DISRT_RAIL, 
           Distance_Roads = DIST_ROAD, WNV_Frequency = FWNVetio) %>% 
    mutate(across(c(Temp_Mean, Temp_Max, Temp_Min), kelvin_to_celsius)) %>% 
    mutate(across(c(Population_Density, Distance_Cities, 
                    Distance_Railroals, Distance_Roads, 
                    WNV_Frequency), as.numeric)) %>% 
    drop_na() %>% 
    mutate(Region = "Afrotropical")

  # Tidy palaearctic dataframe
  pala_clean <- palaearctic %>% 
    select(BIO1, BIO5_MEAN, BIO6_MEAN, BIO12, DENS_POB, 
           DISTCENPOB, DISRT_RAIL, DIST_ROAD, FWNVpal, Region) %>% 
    rename(Temp_Mean = BIO1, Temp_Max = BIO5_MEAN, 
           Temp_Min = BIO6_MEAN,
           Annual_Precipitation = BIO12, Population_Density = DENS_POB, 
           Distance_Cities = DISTCENPOB,
           Distance_Railroals = DISRT_RAIL, 
           Distance_Roads = DIST_ROAD, WNV_Frequency = FWNVpal) %>% 
    mutate(across(c(Temp_Mean, Temp_Max, Temp_Min), kelvin_to_celsius)) %>% 
    mutate(across(c(Population_Density, Distance_Cities, 
                    Distance_Railroals, Distance_Roads, 
                    WNV_Frequency), as.numeric)) %>% 
    drop_na() %>% 
    mutate(Region = "Palaearctic")
  
```

# Part Two: Bar Graphs

``` {r, echo=TRUE, results='hide', message=FALSE, warning=FALSE, fig.width = 12, fig.height = 7, out.width = NULL, dpi = 300}
# ---------------------------
# PART TWO: Bar Graphs
# ---------------------------
  
  # 2.0: The wnv_corr dataframe shows the overall correlation between 
  # WNV frequency and different variables for both regions; it also
  # assigns each value a "Type" ("Ecological" or "Population").
  
  wnv_combined <- bind_rows(afro_clean, pala_clean) 
  
  wnv_corr <- wnv_combined %>% 
    group_by(Region) %>% 
    summarize(
      corr_mean_temp = cor(Temp_Mean, WNV_Frequency, use = "complete.obs"),
      corr_max_temp  = cor(Temp_Max, WNV_Frequency, use = "complete.obs"),
      corr_min_temp  = cor(Temp_Min, WNV_Frequency, use = "complete.obs"),
      corr_precip = cor(Annual_Precipitation, WNV_Frequency, use = "complete.obs"),
      corr_pop = cor(Population_Density, WNV_Frequency, use = "complete.obs"),
      corr_cities = cor(Distance_Cities, WNV_Frequency, use = "complete.obs"),
      corr_roads = cor(Distance_Roads, WNV_Frequency, use = "complete.obs"),
      corr_railroads = cor(Distance_Railroals, WNV_Frequency, use = "complete.obs")) %>% 
      pivot_longer(cols = -Region, names_to = "Variable", values_to = "Correlation") %>% 
      mutate(Type = case_when(
      Variable %in% c("corr_mean_temp", "corr_max_temp", 
                      "corr_min_temp", "corr_precip") ~ "Ecological",
      Variable %in% c("corr_pop", "corr_cities", "corr_roads", 
                      "corr_railroads") ~ "Population")) %>% 
    ungroup()
    
  # 2.1: Using ggplot to create bar graphs showing the correlation between 
  # ecological variables and WNV incidence in afrotropical vs palaearctic regions. 
  
  ggplot(wnv_corr, aes(x = reorder(Variable, Correlation), 
                       y = Correlation, fill = Region)) +
    geom_col(position = position_dodge(width = 0.9)) +
    coord_flip() +
    geom_text(aes(label = round(Correlation, 2), 
                  hjust = ifelse(Correlation >= 0, -0.5, 1.3)), 
                  position = position_dodge(width = 0.9),
                  fontface = "bold",size = 3.5, color = "#5b6482") +
    facet_wrap(~Type, scales = "free_y", 
                  labeller = as_labeller(c("Ecological" = "Ecological Factors",
                                           "Population" = "Population Factors"))) + 
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.2))) +
    scale_fill_manual(values = c("Afrotropical" = "#ffbe5d", 
                                 "Palaearctic" = "#a650a8")) +
    theme_minimal() +
    theme(plot.margin = margin(t = 15, r = 50, b = 15, l = 50),
          plot.title = element_text(hjust = 0.5, face = "bold", size = 16,
                                    margin = margin(b = 10, t = 15)),
          plot.subtitle = element_text(hjust = 0.5, face = "italic",
                                   margin = margin(b = 15), size = 13), 
          axis.title.x = element_text(face = "bold", size = 12,
                                  margin = margin(t = 15)),
          axis.text.y = element_text(face = "bold", color = "black", size = 10), 
          axis.text.x = element_text(face = "bold", color = "black"),
          strip.text = element_text(face = "bold", size = 12, color = "#324ea8"),
          legend.title = element_text(face = "bold", size = 12),
          legend.text = element_text(face = "bold", size = 11),
          legend.position = "bottom") + 
    scale_x_discrete(labels = c("corr_mean_temp" = "Average Temp", 
                                "corr_max_temp" = "Maximum Temp", 
                                "corr_min_temp" = "Minimum Temp", 
                                "corr_precip" = "Annual Precipitation",
                                "corr_pop" = "Population Density", 
                                "corr_cities" = "Distance to Cities",
                                "corr_roads" = "Distance to Roads", 
                                "corr_railroads" = "Distance to Railroads")) +
    labs(title = str_wrap("Correlation of Select Ecological and Population 
                          Demographic Factors on WNV Incidence in AFR and PAL Regions"), 
         subtitle = str_wrap("Environmental drivers of WNV show more regional variation;
                             temperature and precipitation specifically have opposing 
                             correlation patterns between AFR and PAL regions. 
                             Population factors are consistent in both regions; 
                             increasing distance to infrastructure is negatively 
                             correlated, while population density is positively 
                             correlated with WNV incidence.", width = 130),
          x = NULL, 
          y = "Correlation Coefficient")
```

# Part Three: Maps
``` {r, echo=TRUE, results='hide', message=FALSE, warning=FALSE, fig.width = 12, fig.height = 7, out.width = NULL, dpi = 300}
# ---------------------------
# PART THREE: Maps
# ---------------------------
 
  # 3.0: The wnv_avg_corr dataframe shows the overall average correlation between 
  # WNV frequency and all ecological/population variables for both regions. It
  # also calculates Z-score, which will standardize the data for clearer comparison.
  # The wnv_continent_corr dataframe assigns continents to regions so that the data
  # can be mapped using the sf package. 
  
  wnv_avg_corr <- wnv_corr %>%
    mutate(Correlation = abs(Correlation)) %>% 
    group_by(Region, Type) %>%
    summarize(mean_corr = mean(Correlation)) %>% 
    ungroup()
  
  wnv_avg_corr <- wnv_avg_corr %>% 
    mutate(zscore = 
    (mean_corr - (mean(wnv_avg_corr$mean_corr)))/ (sd(wnv_avg_corr$mean_corr)))
  
  wnv_continent_corr <- wnv_avg_corr %>%
    mutate(continent = case_when(
      Region == "Afrotropical" ~ "Africa",
      Region == "Palaearctic" ~ "Europe", 
      TRUE ~ NA_character_))
  
  asia_data <- wnv_avg_corr %>%
    filter(Region == "Palaearctic") %>%
    mutate(continent = "Asia")
  
  wnv_continent_corr <- bind_rows(wnv_continent_corr, asia_data)
    
  # 3.1: Joining wnv_continent_corr with map data.
  
  world <- ne_countries(scale = "medium", returnclass = "sf")
  
  map_data <- left_join(world, wnv_continent_corr, 
                   by = "continent", relationship = "many-to-many") %>% 
                    filter(!is.na(zscore))
  
  # 3.2: Using ggplot to create a map that compares the relative correlation
  # (Z-score) of ecological and population factors on WNV incidence in both
  # regions. 
  
  ggplot(data = map_data) +
    geom_sf(aes(fill = zscore), color = "white", size = 0.1) +
    facet_wrap(~Type, ncol = NULL, 
               labeller = 
                 as_labeller(c("Ecological" = "Ecological Factors",
                               "Population" = "Population Factors"))) +
    coord_sf(xlim = c(-20, 145), ylim = c(-35, 75), expand = FALSE) +
    scale_fill_gradient2(low = "#5097da", 
                         mid = "#e38fc6", 
                         high = "#a83232", 
                         midpoint = 0,
                         limits = c(-1.5, 1.5),
                         breaks = c(-1.5, -0.75, 0, 0.75, 1.5),
                         name = "Relative Correlation\n(Z-score)") +
    theme_void() +
    theme(panel.spacing = unit(1, "lines"),
          plot.margin = margin(r = 10, l = 10),
          plot.title = element_text(hjust = 0.5, face = "bold", size = 17,
                                    margin = margin(b = 5, t = 5)),
          plot.subtitle = element_text(hjust = 0.5, 
                                       face = "italic", size = 15,
                                       margin = margin(b = 5, t = 5)),
          strip.text = element_text(face = "bold", size = 13, 
                                    margin = margin(b = 10, t = 10),
                                    color = "#2b428f"),
          legend.title = element_text(face = "bold", size = 14),
          legend.position = c(0.35, 0.02), 
          legend.direction = "horizontal",           
          legend.text = element_text(size = 8, 
                                     face = "bold", color = "#2b428f")) +
          guides(fill = guide_colorbar(title.position = "top",   
                        title.hjust = 0.5,      
                        label.position = "bottom")) + 
          labs(title = str_wrap("Regional Divergence in Ecological and 
                                Demographic Drivers of WNV Incidence"),
              subtitle = str_wrap("Z-score standardization reveals a divergence 
                                  in WNV drivers across continents by comparing contribution in each
                                  region to the average contribution across all regions 
                                  and variables. Population factors are more influential overall and 
                                  strongest in AFR regions, while ecological 
                                  factors are weaker in comparison but slightly 
                                  stronger in AFR regions. This shows that while population
                                  factors dominate both regions, the intensity of all 
                                  drivers is most pronounced in the AFR region.", width = 130))
```

# Part Four: Pie Charts
``` {r, echo=TRUE, results='hide', message=FALSE, warning=FALSE, fig.width = 12, fig.height = 7, out.width = NULL, dpi = 300}
# ---------------------------
# PART FOUR: Pie Charts
# ---------------------------
  
  # 4.0: The pie_normalized dataframe calculates the estimated percent contribution 
  # of each variable by squaring the correlation value (to make positive) 
  # and dividing that number by the sum of all contributing variables for each
  # contributing factor. The combined_pie dataframe summarizes the percent
  # contribution data across each region for the "both" pie chart. 
  
  pie_normalized <- wnv_corr %>%
    mutate(Contribution = Correlation^2) %>% 
    group_by(Region) %>%
    mutate(Percentage = 
           Contribution / sum(Contribution, na.rm = TRUE) * 100) %>%
    ungroup() 
  
  combined_pie <- pie_normalized %>%
    group_by(Variable) %>%
    summarize(Percentage = mean(Percentage)) %>%
    mutate(Region = "Overall Average")  
  
  final_pie_data <- bind_rows(pie_normalized, combined_pie)
  
  variable_order <- c("corr_mean_temp","corr_min_temp", 
                      "corr_max_temp", "corr_precip",
                      "corr_pop", "corr_cities", "corr_roads",
                      "corr_railroads")
  
  final_pie_data$Variable <- factor(final_pie_data$Variable, 
                                    levels = variable_order)
  
  # 4.1: Using ggplot to create pie charts that compare the estimated 
  # contribution of each ecological and population factor on WNV incidence
  # in each region and across both regions. 
  
  ggplot(final_pie_data, aes(x = "", y = Percentage, fill = Variable)) +
    geom_col(width = 1, color = "white") +
    coord_polar("y", start = 0) +
    facet_wrap(~Region) + 
    scale_fill_manual(
      values = c("corr_mean_temp" = "#03440C", 
                "corr_max_temp" = "#32A86F", 
                "corr_min_temp" = "#13505B", 
                "corr_precip" = "#669BBC",
                "corr_pop" = "#C266C7", 
                "corr_cities" = "#FA57A9",
                "corr_roads" = "#832F96", 
                "corr_railroads" = "#E595E0"),
      labels = c(
        "corr_mean_temp" = "Average Temp", 
        "corr_max_temp" = "Maximum Temp", 
        "corr_min_temp" = "Minimum Temp", 
        "corr_precip" = "Annual Precipitation",
        "corr_pop" = "Population Density", 
        "corr_cities" = "Distance to Cities",
        "corr_roads" = "Distance to Roads", 
        "corr_railroads" = "Distance to Railroads")) +
    theme_void() + 
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 17,
                                    margin = margin(b = 2, t = 1)),
          plot.subtitle = element_text(hjust = 0.5, face = "italic",
                                       size = 15,
                                       margin = margin(b = 10, t = 5)),
          legend.title = element_text(hjust = 0, face = "bold", size = 12),
          plot.margin = margin(r = 10, l = 10),
          legend.position = "left",
          legend.text = element_text(size = 11, face = "bold", 
                                     color = "#2b428f", 
                                     margin = margin(l = 1, t = 1)),
          strip.text = element_text(face = "bold", size = 14, 
                                    color = "#324ea8", 
                                    margin = margin(t = 5, b = 1))) +
          labs(title = str_wrap("Estimated Percent Contribution of Ecological 
                                & Population 
                                Factors to WNV Incidence"), 
               subtitle = str_wrap("Population factors, primarily distance to 
                                    infrastructure, have the highest contribution 
                                    to WNV incidence across all regions, showing their 
                                    greatest relative contribution in the PAL region,
                                    despite overall contribution being higher in AFR.
                                    Ecological factors are more significant in AFR regions, 
                                    particularly maximum temperature and annual 
                                    precipitation, which represent a larger proportion of 
                                    the local drivers than in PAL.", width = 100),
                fill = "Ecological and Population\nFactors")
```

\pagebreak
# References

CDC. (2024, May 17). West Nile: Symptoms, Diagnosis, & Treatment. West Nile Virus. https://www.cdc.gov/west-nile-virus/symptoms-diagnosis-treatment/index.html<br>

José‐María García‐Carrasco, Souviron‐Priego, L., Antonio‐Román Muñoz, Olivero, J., Fa, J. E., & Real, R. (2024). Present and future situation of West Nile virus in the Afro‐Palaearctic pathogeographic system. Ecography. https://doi.org/10.1111/ecog.06941

